RILS-ROLS: Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares

### 2022-10-22 doc/paper.tex
Update in Section 3.3.1

 \subsection{Ordinary least square method}
   Ordinary least square method (OLS) is a linear regression technique. It is based on applying the least-square method to minimize the square residual (error) sum  between actual and predicted values (given by the model). More precisely, given data $({X}=(X_1, \ldots, X_d)^T, y)$, the task is to determine linear mapping $\tilde{y} = k \textbf{x} + b$, that is coefficients (line slope) $k = (k_1, \ldots, k_d)$ and $b$ (intercept), so that $ \sum_{i} |\tilde{y}_i - y_i|^2 $ is minimized. This sum is also know as the sum of squared error (SSE). There are many methods to minimize SSE. One of the analytical approaches is the calculus-oriented -- it takes into account   partial derivation w.r.t. $k$ of the cost function $ \sum_{i} |\tilde{y}_i - y_i|^2 $, getting

@@

RILS-ROLS algorithms receives training dataset $D_{tr}$ as input. $D_{tr}$ can have very large number of records, so the first step of RILS-ROLS is to make a random sample $D_{tr}'$ of size $sample_{size}$. Initial sample size is calculated as $\max(0.01 \cdot n_{tr}, 100)$. Sample is later dynamically adjusted -- when there are no solution improvements for some number of iterations, the sample size is doubled (lines~\ref{line:sampleAdj1}-\ref{line:sampleAdj2} of Algorithm~\ref{alg:rilsrols}).

As previously stated, solution is usually represented as a tree. We use simple solution initialization -- tree root node is set to zero constant. We interchangeably use two solution variables: i) $s$ denotes starting (or working) solution and ii) $bs$ stands for the best solution so far. Solution quality is measured by evaluating fitness function (more on this in the following Section~\ref{sec:fitness}). Before entering the main loop, the best solution is set to initial solution (line~\ref{line:solSet}).

Main loop iterates as long as none of termination criteria is met: i) maximal running time has been reached; ii) maximal number of fitness calls has been made; iii) best solution is sufficiently good.
The third criterion is based on the $R^2$ and $RMSE$ of the best solution. If $R^2$ is sufficiently close to 1 and, at the same time, $RMSE$ is sufficiently close to 0, the algorithm stops prematurely -- this significantly reduces the running times for the majority of tested instances.

On of the first steps in the main loop is to generate perturbations near the starting solution $s$ (line~\ref{line:sPert}). 
As the name of this procedure (\texttt{All1Perturbations}) suggests, perturbation is local -- meaning the closeness of starting solution $s$ and any of perturbations is 1 (we call it 1-perturbation sometimes). The precise way of generating perturbation is described in a separate Section~\ref{sec:pertGen}.

Candidate perturbations are being improved by performing OLS coefficient fitting (procedure \texttt{FitOLS}). This means that coefficients in linear combinations are being set by means of ordinary least squares method.  
After this step, perturbations are usually better suited to given sample data $D_{tr}'$. 
Further, they are sorted w.r.t. $R^2$ metric in the descending order (line~\ref{line:orderR2}).

Now, the algorithm enters internal loop -- it iterates over ordered perturbations and aims to find the one which improves the best solution $bs$. But, before comparing candidate perturbation solution $p$ with $bs$, $p$ is first simplified (line~\ref{line:simp}) after which the local search is performed (line~\ref{line:ls}).
Solution simplification is done in a symbolical fashion by popular Python library called SymPy~\cite{sympy}.
Local search tries to find local optima expressions close to the given $p$ -- there is a separate Section~\ref{sec:ls} dedicated to local search. 
Finally, fitness function value of $p$ is compared to fitness function value of $bs$. If it is better, $bs$ is updated correspondingly and internal loop that goes across ordered perturbations is immediately stopped (\emph{first-improvement} strategy). Otherwise, the next perturbation is probed. 
Note that probed perturbations are remembered in a set denoted by $perturbations_{tried}$. The goal is to avoid checking the same perturbation multiple times (lines \ref{line:avoid1}-\ref{line:avoid2}).

If some of $s_{perturbations}$ around starting solution $s$ yielded an improvement, $bs$ becomes the starting solution $s$ in the next iteration of the main loop (line~\ref{line:improved}). 
Otherwise, it makes no sense to set starting solution to $bs$, since nothing changed -- search is \emph{trapped} in local optimum. Therefore, a randomness needs to be introduced. First, a set of local perturbations around $bs$ is generated ($start_{candidates}$) in the same manner as before (procedure \texttt{All1Perturbations}). If at least one of these was not previously used as a starting solution ($start_{tried}$), a single perturbation from the $start_{candidates} \setminus start_{tried}$ is randomly picked (line~\ref{line:randPick}). There is small chance that $start_{candidates} \setminus start_{tried} = \emptyset$. When that happens, the set of starting solution candidates is equal to perturbations of some randomly selected perturbation of $bs$ (lines \ref{line:pert21}-\ref{line:pert22}) -- which effectively means that perturbations on distance 2 from $bs$ are used. There is a very small chance that these 2-perturbations will have empty set difference with $start_{tried}$ set of starting solutions. Therefore, 3-perturbations are not considered.

@@

Symbolic regression objective is to determine the expression governing available data. It is also allowed to obtain some equivalent expression, having in mind there are many ways to express some symbolical equation. Although logically sound and intuitive, this objective is not quantifiable during the solution search/training phase, simply because the goal expression is not known at that point -- only target values for some of the input features are known. So, different numerical metrics are being used in literature to guide the symbolic regression search process. The most popular are the coefficient of determination, also known as $R^2$ and mean squared error (MSE) or root mean squared error (RMSE). The important aspect of the solution quality might also be the solution expression complexity. This follows the Occam's razor principle that simpler solution is more likely to be a correct one. 
The search process of RILS-ROLS is guided by the non-linear combination of $R^2$, RMSE and solution expression size (complexity) presented in Equation~\ref{eq:fitness}.

@@

Since the presented fitness function needs to be minimized, the following conclusions may be drawn:
\begin{itemize}
	\item higher $R^2$ is preferred -- ideally, when $R^2(s)=1$, the effect of term $1+R^2(s)$ is neutralized, since 1 is neutral for multiplication;
	\item lower RMSE is preferred, ideally, when $RMSE(s)=0$, the whole term $(1+RMSE(s))$ becomes 1;
	\item since $penalty_{size} > 0$, larger expressions tend to have higher fitness (which follows the Occam's razor principle).
\end{itemize}

>>>>

 \subsection{Ordinary least square method}\label{sec:ols}
   Ordinary least square method (OLS) is a linear regression technique. It is based on applying the least-square method to minimize the square residual (error) sum  between actual and predicted values (given by the model). More precisely, given data $({X}=(X_1, \ldots, X_d)^T, y)$, the task is to determine linear mapping $\tilde{y} = k \textbf{x} + b$, that is coefficients (line slope) $k = (k_1, \ldots, k_d)$ and $b$ (intercept), so that $ \sum_{i} |\tilde{y}_i - y_i|^2 $ is minimized. This sum is also know as the sum of squared error (SSE). There are many methods to minimize SSE. One of the analytical approaches is the calculus-oriented -- it takes into account   partial derivation w.r.t. $k$ of the cost function $ \sum_{i} |\tilde{y}_i - y_i|^2 $, getting

@@

RILS-ROLS algorithms receives training dataset $D_{tr}$ as input. $D_{tr}$ can have very large number of records, so the first step of RILS-ROLS is to choose a random sample $D_{tr}' \subseteq D_{tr}$ of size $sample_{size}$. Initial the sample size is calculated as $\max(0.01 \cdot n_{tr}, 100)$. The size of sample is later dynamically adjusted through the algorithm's iterations -- when there are no solution improvements for some number of iterations, the sample size is doubled (lines~\ref{line:sampleAdj1}-\ref{line:sampleAdj2} of Algorithm~\ref{alg:rilsrols}).

As previously stated, solution is usually represented by means of a tree. We use simple solution initialization -- tree root node is set to zero constant. We interchangeably use two solution variables: ($i$) $s$ denotes starting (or working) solution and ($ii$) $bs$ stands for the best solution so far. Solution quality is measured by evaluating the fitness function (more about it in the subsequent Section~\ref{sec:fitness}). Before entering the main loop, the best solution \emph{bs} is set to the initial solution (line~\ref{line:solSet}).

Main loop iterates as long as none of termination criteria is met: ($i$) maximal running time has been reached; ($ii$) maximal number of fitness calculations has been made; ($iii$) best solution is sufficiently good, 
w.r.t. the $R^2$ and $RMSE$ of the best solution. More precisely, if $R^2$ is sufficiently close to 1 and, at the same time, $RMSE$ is sufficiently close to 0, the algorithm stops prematurely -- this significantly reduces the running times for the majority of tested instances.

One of the first steps in the main loop is to generate perturbations near the starting solution $s$ (line~\ref{line:sPert}). 
As the name of this procedure (\texttt{All1Perturbations}) suggests, perturbation step is local -- meaning the closeness of starting solution $s$ and any of perturbations is 1 (we call it 1-perturbation sometimes). The precise way of generating perturbation is described separately, in Section~\ref{sec:pertGen}.

Candidate perturbations are being improved by performing OLS coefficient fitting (procedure \texttt{FitOLS}). This means that coefficients in any of linear combinations of current solution are being set by means of ordinary least squares method, described in Section~\ref{sec:ols}.   
After this step, perturbations are usually better suited to given sample data $D_{tr}'$. 
Further, these perturbations are sorted w.r.t. $R^2$ metric in the descending order (line~\ref{line:orderR2}).

Now, the algorithm enters the internal loop -- it iterates over the ordered perturbations and aims to find the one which improves the best solution $bs$. But, before comparing candidate perturbation solution $p$ with $bs$, $p$ is first simplified (line~\ref{line:simp}) after which the local search is performed (line~\ref{line:ls}).
Solution simplification is done in a symbolical fashion by popular Python library called SymPy~\cite{sympy}.
Local search tries to find local optima expressions close to the given $p$ -- explained in details in Section~\ref{sec:ls}.  
Finally, fitness function value of $p$ is compared to fitness function value of $bs$. If a new incumbent is found, $bs$ is updated correspondingly and internal loop that goes across ordered perturbations is immediately terminated (it works in a  \emph{first-improvement} strategy). Otherwise, the next perturbation is probed. 
Note that the probed perturbations are stored in a set denoted by $perturbations_{tried}$. The goal is to avoid checking the same perturbation multiple times (lines \ref{line:avoid1}-\ref{line:avoid2}), i.e. this structure serves as a kind of tabu list, initially proposed in the famous Tablu search algorithm, see~\cite{glover1998tabu}.

If some of $s_{perturbations}$ around starting solution $s$ yielded an improvement, $bs$ becomes the starting solution $s$ in the next iteration of the main loop (line~\ref{line:improved}). 
Otherwise, it makes no sense to set starting solution to $bs$, as nothing changed -- the search is being  \emph{trapped} in a local optimum. In order to avoid this undesired situation, a randomness needs to be triggered. First, a set of local perturbations around $bs$ is generated ($start_{candidates}$) in the same manner as before (procedure \texttt{All1Perturbations}). If at least one of these was not previously used as a starting solution ($start_{tried}$), a single perturbation from the $start_{candidates} \setminus start_{tried}$ is randomly picked (line~\ref{line:randPick}). There is a insignificant chance that $start_{candidates} \setminus start_{tried} = \emptyset$. When that happens, the set of starting solution candidates is equal to perturbations of some randomly selected perturbation of $bs$ (lines \ref{line:pert21}-\ref{line:pert22}) -- which effectively means that the perturbations with distance 2 from $bs$ are used. Indeed, there is a minor chance that these 2-perturbations will have empty set difference with $start_{tried}$ set of starting solutions. Therefore, more expensive 3-perturbations are not considered in our algorithm.

@@

Symbolic regression objective is to determine the expression governing available data. It is also allowed to obtain some equivalent expression, keeping in mind that there are multiple ways to express some symbolical equation. Although logically sound and intuitive, this objective is not quantifiable during the solution search/training phase, simply because the goal expression is not known at that point as only target values for some of the input features are known. Thus, different numerical metrics are being used in literature to guide the symbolic regression search process. The most popular are the coefficient of determination, also known as $R^2$ and mean squared error (MSE) or root mean squared error (RMSE). The important aspect of the solution quality might also be the solution expression complexity. This follows the Occam's razor principle~\cite{costa2020fast}~ that simpler solution  is more likely to be a correct one. 
The search process of RILS-ROLS is guided by the non-linear combination of $R^2$, RMSE and solution expression size (complexity) presented in Equation~\ref{eq:fitness}.

@@

Since the presented fitness function needs to be minimized, the following conclusions may be drawn:
\begin{itemize}
	\item higher $R^2$ is preferred -- ideally, when $R^2(s)=1$, the effect of term $1+R^2(s)$ is neutralized; %since 1 is neutral for multiplication;
	\item lower RMSE is preferred, ideally, when $RMSE(s)=0$, the whole term $(1+RMSE(s))$ becomes 1;
	\item since $penalty_{size} > 0$, larger expressions tend to have higher fitness (which follows the Occam's razor principle); therefore, we award solutions that are more simple with its representation. 
\end{itemize}
